\chapter*{Abstract}
\addcontentsline{toc}{prelim}{ABSTRACT}

The theory of efficient markets states that capital markets
instantly adjust to incorporate all relevant information. 
When new information arrives about a firm or an industry,
efficent markets theory dictates that the price of the firm
or industry in question move to incorporate this
information.  This theory provides a basis for studying the
effects of news on firms.  By performing what is called an
event study, a researcher can assess how the market
perceives some news on the company in question.

The study here focuses on the effect of the explosion on
Tuesday, January 28, 1986, of the space shuttle {\em
Challenger}.  The null hypothesis that was disproved was
that due to the explosion, firms involved in the shuttle
program would expect to see a drop in returns on the day of
the accident.  This is what was seen, except for one
company, Morton Thiokol, the manufacturer of the solid
rocket booster.  The cause of the accident was completely
unknown until a few days later, but the market reacted on
the day of the accident; Morton Thiokol's firm value dropped
twelve percent.  The reaction largely stemmed from knowledge
that the joint design of the solid rocket boosters was
faulty and nothing had been done about it.

The market expected substantially lower returns in the
future for Morton Thiokol.
Morton Thiokol dropped out of the bidding for
NASA's next generation solid rocket boosters, some believe
because of political pressure.
The contract had to receive Congress's approval, but after
the revelations about the faulty design of the solid rocket
boosters joint, many felt it unlikely that Morton Thiokol
would ever win the contract.
Morton Thiokol also performed {\em pro bono} the design work
on the new solid rocket boosters.

On the day of the accident and because of the faulty design
of the solid rocket booster's joints, the market indicated
that Morton Thiokol would face some future loss.
The explosion caused information of the faulty joint to
become widely held, and the market reacted to past information
it held about substantial penalties paid by at-fault
companies.
