\chapter{Conclusion}

Chapter~\ref{litrev} reviewed the relevant efficient market, capital asset pricing, and event study literature in order to provide a foundation for the study that followed. The joint-hypothesis problem was reviewed as were problems in performing event studies. The joint-hypothesis problem is caused by the requirement of an asset-pricing model to accompany any tests of market efficiency. This confounds results of studies of efficiency in markets because the result could be due to an inappropriate asset pricing model. This is one of the reasons that Fama refers to various tests for degrees of market efficiency rather than tests only for the existence of efficiency in security markets.

Chapter~\ref{history} reviews the space shuttle program and the decision making that went into the design of today's space shuttle. The space shuttle program was conceived in the late 1960s as a means of transporting satellites and people cheaply into space. The development costs of the ideal (ideal in the sense of low operating costs) system were too high, and Congress and the Office of Management and Budget balked. This caused NASA to redesign the system to use, among other things, solid rocket technology. Although solid rockets had never before been used in manned space flight, they had a proven record in missiles and satellite launches. This introduced a concern into the program because the crew would be unable to escape the shuttle during the two minute period when the solid rocket boosters burn.

The chapter then turned to the accident and analyzed the accident from the viewpoint of telemetry and enhanced visual images. This showed that the right solid rocket booster caused the accident, but this was not obvious on the day of the accident. In fact, on the day of the accident it appeared that the external tank and the main orbiter engines were the cause of the accident. This was later ruled out by the Rogers Commission, but the information about suspected causes leaked out earlier after some of the enhanced images were released by NASA.

The Commission eventually found that the joint on the solid rocket boosters was faulty, and that this problem was known for some time. The remainder of the chapter details the length of time this knowledge was known and by whom. The problem surfaced in about 1977 during the pre-launch days of the shuttle. It was in 1985 that the problem became more evident because of a method NASA used to assure that the O-rings between joints of the solid rocket boosters were sealed. The test they used blew holes in the putty that guarded the O-rings against the hot gasses of combustion as the central core of the SRB burned. The blow holes allowed hot gasses to reach the O-rings and burn (or erode) them.

Chapter~\ref{empiric} discussed the statistical results of this study, finding that the market has quite a depth of information available about causes. The results showed what was expected; all firms experienced negative returns. The finding that lends credence to the hypothesis of efficient markets is that the return on Morton Thiokol was negative twelve percent on the day of the accident when no one really knew what caused the accident, but many suspected the O-rings because it had been a problem for some time. The market thus performed to expectations by instantly incorporating information about the cause of the {\em Challenger} accident into the firm at fault.